Appendix: Why the complement of AUC isn't a casewise CV criterion

### 2024-07-08 scratch/notes_georges/JSS-article/JSS-article-3-GM.Rmd
Contrasting cluster-based and case-based CV down to 4 pages.

The \code{Auto} dataset contains information about 392 cars:

@@

The four datasets differ in the relative magnitude of between-cluster variance compared with within-cluster variance. They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.

@@

Our four contrived data sets (see below) illustrate possible results for data obtained in such a scenario. The relative configuration of dosages `x` and symptoms `y` are identical within patients in each of the four datasets. Within patients, higher dosages are generally associated with a reduction in symptoms.

Between patients, however, higher dosages are associated with higher levels of symptoms. A plausible mechanism is a reversal of causality: Within patients, higher dosages alleviate symptoms, but, between patients, higher morbidity causes the prescription of higher dosages.

The four data sets differ in the between-patient variance of patient centroids from a common between-patients regression line. The data sets exhibit a progression from low to high variance around the common regression line.

<!-- We start by generating a data set to serve as a common template for the four sample data sets.  We then apply different multipliers to the between-patient variation using parameters consistent with the description above to highlight the issues that arise in cross-validating mixed-effects models:[^data-template] -->

<!-- [^data-template]: We invite the interested reader to experiment with varying the parameters of our example.  -->

@@

Data_template <- expand.grid(patient = 1:Nb, obs = 1:Nw) |>
  within({
    xw <- seq(-2, 2, length.out = Nw)[obs]
    x <- patient + xw
    xm  <- ave(x, patient)   # within-patient mean
    # Scaled random error within each SD_ratio_i group
    re_std <- scale(resid(lm(rnorm(Nb*Nw) ~ x)))
    re_between <- ave(re_std, patient)
    re_within <- re_std - re_between
    re_between <- scale(re_between)/sqrt(Nw)
    re_within <- scale(re_within)
  })
Data <- do.call(
  rbind,
  lapply(
    1:Nv,
    function(i) {
      cbind(Data_template, SD_ratio_i = i)
    }
  )
)
Data <- within(
  Data,
  {
    SD_within_ <- SD_within[SD_ratio_i]
    SD_between_ <- SD_between[SD_ratio_i]
    SD_ratio <- SD_ratio[SD_ratio_i]
    y <- 10 +
      Bb * xm +                  # contextual effect
      Bw * (x - xm) +            # within-patient effect
      SD_within_ * re_within +   # within patient random effect
      SD_between_ * re_between   # adjustment to between patient random effect
  }
)
```
Figure \ref{fig:plot1} displays scatterplots of each data set showing estimated 50% concentration ellipses for each cluster.[^lattice]

@@

```{r plot1, echo=FALSE}
#| out.width = "100%",
#| fig.height = 3.5,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD. The ellipses are estimated 50% concentration ellipses for each patient. The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
library("lattice")
library("latticeExtra")
layer <- latticeExtra::layer # to avert conflicts with ggplot2::layer
xyplot(y ~ x | SD_ratio, data = Data, group = patient,
       par.strip.text = list(cex = 0.7),
       scales = list(x = list(alternating = FALSE)),
       layout = c(Nv, 1),
       par.settings = list(superpose.symbol = list(pch = 1, cex = 0.7))) +
  layer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                      level = 0.5),
        panel.abline(a = 10, b = 1))
```
The population from which these datasets are generated has a between-patient effect of dosage of 1 and a within-patient effect of $-0.5$. The researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models have sequentially nested fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`. The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`. We thus fit four models to four datasets for a total of 16 models.

```{r model-fits,include=FALSE}
model.formulas <- c(
  ' ~ 1'             =  y ~ 1 + (1 | patient),
  '~ 1 + x'          =  y ~ 1 + x + (1 | patient),
  '~ 1 + x + xm'     =  y ~ 1 + x + xm + (1 | patient),
  '~ 1 + I(x - xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
)
fits <- lapply(split(Data, ~ SD_ratio),
               function(d) {
                 lapply(model.formulas, function(form) {
                   glmmTMB(form, data = d)
                 })
               })
```
We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or "BLUPs"---as would be used for cross-validation based on cases (i.e., occasions within patients):
```{r predict,include=FALSE}
pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  
pred.BLUPs <- lapply(fits, lapply, predict)
```
<!-- We then prepare the data for plotting: -->

```{r data-predictions,include=FALSE}
Dataf <- lapply(split(Data, ~ SD_ratio),
                function(d) {
                  lapply(names(model.formulas), 
                         function(form) cbind(d, formula = form))
                }) |> 
  lapply(function(dlist) do.call(rbind, dlist)) |> 
  do.call(rbind, args = _) |> 
  within(
    {
      pred.fixed <- unlist(pred.fixed)
      pred.BLUPs <- unlist(pred.BLUPs)
      panel <- factor(formula, levels = c(names(model.formulas), 'data'))
    }
  )
Data$panel <- factor('data', levels = c(names(model.formulas), 'data'))
```
The fixed-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-fixed} along with a summary of the data using 50% concentration ellipses.  Figure \ref{fig:plot-fits-blups} similarly displays predicted BLUPs. 
```{r plot-fits-fixed,echo=FALSE}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed-effect predictions using each model applied to data sets with varying between- and within-patient variance ratios.  The top row shows summaries of the within-patient data using estimated 50% concentration ellipses."
{
  xyplot(y ~ x |SD_ratio * panel, Data,
         groups = patient, type = 'n',    
         par.strip.text = list(cex = 0.6),
         scales = list(x = list(alternating = FALSE)),
         drop.unused.levels = FALSE,
         ylab = list('fixed-effect predictions',cex=.7)) + 
    glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                         level = 0.5),
           panel.abline(a = 10, b = 1)) +
    xyplot(pred.fixed  ~ x |SD_ratio * panel, Dataf, type = 'l', 
           groups = patient,
           drop.unused.levels = F)
}|> useOuterStrips()
```
<!-- The BLUPs from these models, shown in Figure \ref{fig:plot-fits-blups}, are produced as follows: -->
```{r plot-fits-blups,echo=FALSE}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios. The top row shows summaries of the within-patient data using estimated 50% concentration ellipses."
{   xyplot(y ~ x | SD_ratio * panel, Data,
         groups = patient, type = 'n',
         drop.unused.levels = F, 
         par.strip.text = list(cex = 0.6),
         scales = list(x = list(alternating = FALSE)),
         ylab = list('fixed- and random-effect predictions (BLUPS)', cex = 0.7)) +
    glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                         level = 0.5),
           panel.abline(a = 10, b = 1)) +
    xyplot(pred.BLUPs  ~ x | SD_ratio * panel, Dataf, type = 'l', 
           groups = patient,
           drop.unused.levels = F) 
}|> useOuterStrips()
```
Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`. The inclusion of a contextual variable in the model corrects this problem.

Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform with the within-patient slopes for the `~ 1 + x` model in the two datasets with the smallest between-patient variances.

For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for datasets with larger between-subject variance.  With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.

@@

We will discover that case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model for the '5 / 2.5' dataset, but cluster-based cross-validation prefers the latter model to the former. The situation is entirely reversed with the '8 / 2.5' dataset.

@@

We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only. In order to reduce between-model random variability in comparisons of models on the same dataset, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.

```{r echo=FALSE,include=FALSE}
library(cv) # unclear why it's necessary to reload cv
.opts <- options(warn = -2) 
```
```{r cross-validation,cache=TRUE,message=FALSE,include=FALSE}
model_lists <- lapply(fits, function(fitlist) do.call(models, fitlist))
cvs_cases <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]], k = 10, 
              data = split(Data, ~ SD_ratio)[[i]])
         })
cvs_clusters <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]],  k = 10, 
              data = split(Data, ~SD_ratio)[[i]], 
              clusterVariables = 'patient')
         })
```
For a given dataset, we can plot the results for a list of models using the plot method for `"cvList"` objects. 
For example, a comparison cluster-based cross-validation using fixed effects for the models fit to the third dataset (with "SD ratio = 6 / 2.5") is shown in Figure \ref{fig:plot-cv-example}.
```{r plot-cv-example,echo=FALSE}
#| out.width = "60%",
#| fig.height = 3.8,
#| fig.cap = "10-fold cluster-based cross-validation comparing random intercept mixed models with varying fixed effects."
plot(cvs_clusters[[3]], main="Comparison of Fixed Effects",)
```
The results for all datasets, using both cluster-based and case-based cross-validation, are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
```{r cross-validation-data,include=FALSE}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })
dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r cross-validation-data-plot,echo=FALSE}
#| out.width = "100%",
#| fig.height = 3.3,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(clusters + cases ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', reverse.rows = T, columns = 2, cex = 0.7), 
       type = 'b', ylab = list('CV criterion (MSE)', cex = .7),
       xlab = '',
       par.strip.text = list(cex = 0.7),
       layout= c(Nv, 1),
       par.settings =
         list(superpose.line=list(lty = c(1, 2), lwd = 1),
              superpose.symbol=list(cex = 0.7)),
       scales = list(y = list(log = TRUE, cex = 0.7), 
                     x = list(alternating = F, rot = 45, cex = 0.7))) 
```
In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. With all our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

@@

The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
```{r Auto-transformed-scatterplot-matrix}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data"
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread=FALSE), data=A, pch=".")
```
And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and Figure \ref{fig:Auto-CR-plots-transformed}):
```{r Auto-CR-plots-transformed}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Component+residual plots for the model fit to the transformed \\code{Auto} data"
crPlots(m, pch=".")
```

After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
```{r}
library("MASS")
m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
summary(m.step)
```
The selected model includes three of the numeric predictors, \code{horsepower}, \code{weight}, and \code{acceleration}, along with the factors \code{year} and \code{origin}. We can calculate the MSE for this model, but we expect that the result will be optimistic because we used the whole data to help specify the model:
```{r MSE-whole-selected-model}
mse(Auto$mpg, exp(fitted(m.step)))
```
This is considerably smaller than the MSE for the original working model:
```{r MSE-working-model}
mse(Auto$mpg, fitted(m.auto))
```
A perhaps subtle point is that we compute the MSE for the selected model on the original \code{mpg} response scale rather than the log scale, so as to make the selected model comparable to the working model.[^medAbsErr]

@@

compareFolds(cvs)
```
The other arguments to \code{cv()} are (see \code{?cv:::cv.function} for additional optional arguments and details):

>>>>

The \code{Auto} data set contains information about 392 cars:

@@

The four data sets differ in the relative magnitude of between-cluster variance compared with within-cluster variance. They serve to illustrate how fitting mixed models, and, consequently, the cross-validation of mixed models, is sensitive to relative variance, which affects the degree of shrinkage of within-cluster estimates of effects towards between-cluster estimates.

@@

Within patients, higher dosages are generally associated with a reduction in symptoms. Between patients, however, higher dosages are associated with higher levels of symptoms. A plausible mechanism is a reversal of causality: Within patients, higher dosages alleviate symptoms, but, between patients, higher morbidity causes the prescription of higher dosages.

To construct the four data sets, we first construct a template data set from a population with a between patient effect of dosage of $1.0$ and a within-patient effect of $-0.5$. We then apply four different multipliers to the between-patient variation from the regression line of patient centroids. The resulting data sets, shown in Figure \ref{fig:plot1}, differ in the between-patient variance of patient centroids from a common between-patient regression line, exhibiting a progression from low to high variance around the common regression line.[^lattice]

Note that each patient has the same relative configuration of dosages and symptoms in each of the four data sets. To help visualize the structure of the data, estimated 50% concentration ellipses are shown for each patient.

Using such data, the researcher may attempt to obtain an estimate of the within-patient effect of dosage through the use of mixed models with a random intercept. We will illustrate how this approach results in estimates that are highly sensitive to the relative variance at the two levels of the mixed model by considering four models, all of which include a random intercept.  The first three models are sequentially nested with fixed effects: (1)`~ 1`, intercept only; (2)`~ 1 + x`, intercept and effect of `x`; and (3) `~ 1 + x + xm`, intercept, effect of `x`, and a contextual variable, `xm`, consisting of the within-patient mean of `x`. The fourth model, `~ 1 + I(x - xm)`, uses an intercept and the centered-within-group variable `x - xm`. We thus fit four models to four data sets for a total of 16 models.

@@

Data_template <- expand.grid(patient = 1:Nb, obs = 1:Nw) |>
  within({
    xw <- seq(-2, 2, length.out = Nw)[obs]
    x <- patient + xw
    xm  <- ave(x, patient)   # within-patient mean
    # Scaled random error within each SD_ratio_i group
    re_std <- scale(resid(lm(rnorm(Nb*Nw) ~ x)))
    re_between <- ave(re_std, patient)
    re_within <- re_std - re_between
    re_between <- scale(re_between)/sqrt(Nw)
    re_within <- scale(re_within)
  })
Data <- do.call(
  rbind,
  lapply(
    1:Nv,
    function(i) {
      cbind(Data_template, SD_ratio_i = i)
    }
  )
)
Data <- within(
  Data,
  {
    SD_within_ <- SD_within[SD_ratio_i]
    SD_between_ <- SD_between[SD_ratio_i]
    SD_ratio <- SD_ratio[SD_ratio_i]
    y <- 10 +
      Bb * xm +                  # contextual effect
      Bw * (x - xm) +            # within-patient effect
      SD_within_ * re_within +   # within patient random effect
      SD_between_ * re_between   # adjustment to between patient random effect
  }
)
```

@@

```{r plot1, echo=FALSE}
#| out.width = "100%",
#| fig.height = 3.0,
#| fig.cap = "Data sets showing identical within-patient configurations with increasing between-patient variance. The labels above each panel show the between-patient SD/within-patient SD. The ellipses are estimated 50% concentration ellipses for each patient. The population between-patient regression line, $E(y) = 10 + x$, is shown in each panel."
library("lattice")
library("latticeExtra")
layer <- latticeExtra::layer # to avert conflicts with ggplot2::layer
xyplot(y ~ x | SD_ratio, data = Data, group = patient,
       ylab = list('symptoms', cex = 0.7),
       xlab = list('dosage', cex = 0.7),
       par.strip.text = list(cex = 0.7),
       scales = list(cex = 0.7, x = list(alternating = FALSE)),
       layout = c(Nv, 1),
       par.settings = list(superpose.symbol = list(pch = 1, cex = 0.7))) +
  layer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,  
                      level = 0.5),
        panel.abline(a = 10, b = 1))
```

```{r model-fits,include=FALSE}
model.formulas <- c(
  ' ~ 1'             =  y ~ 1 + (1 | patient),
  '~ 1 + x'          =  y ~ 1 + x + (1 | patient),
  '~ 1 + x + xm'     =  y ~ 1 + x + xm + (1 | patient),
  '~ 1 + I(x - xm)'  =  y ~ 1 + I(x - xm) + (1 | patient)
)
fits <- lapply(split(Data, ~ SD_ratio),
               function(d) {
                 lapply(model.formulas, function(form) {
                   glmmTMB(form, data = d)
                 })
               })
```
We proceed to obtain predictions from each model based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., patients), and for fixed and random effects---so-called best linear unbiased predictions or "BLUPs"---as would be used for cross-validation based on cases (i.e., occasions within patients):
```{r predict,include=FALSE}
pred.fixed <- lapply(fits, lapply, predict, re.form = ~0)  
pred.BLUPs <- lapply(fits, lapply, predict)
```
```{r data-predictions,include=FALSE}
Dataf <- lapply(split(Data, ~ SD_ratio),
                function(d) {
                  lapply(names(model.formulas), 
                         function(form) cbind(d, formula = form))
                }) |> 
  lapply(function(dlist) do.call(rbind, dlist)) |> 
  do.call(rbind, args = _) |> 
  within(
    {
      pred.fixed <- unlist(pred.fixed)
      pred.BLUPs <- unlist(pred.BLUPs)
      panel <- factor(formula, levels = c(names(model.formulas), 'data'))
    }
  )
Datap <- reshape(Dataf, direction = "long", sep = '.',
                 varying = c('pred.fixed', 'pred.BLUPs'),
                 timevar = 'prediction_type')
Datap$panel2 <- with(Datap, paste0(formula,":",prediction_type))
flevels <- c(unique(Datap$panel2), 'data')
Datap$panel2 <- factor(Datap$panel2, levels = flevels)
Data$panel <- factor('data', levels = c(names(model.formulas), 'data'))
Data$panel2 <- factor('data', levels = flevels)
```
The fixed- and random-effects predictions from these models are displayed in Figure \ref{fig:plot-fits-combined} along with a summary of the data using 50% concentration ellipses.

```{r plot-fits-combined,echo=FALSE}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Fixed- and random-effect predictions (BLUPs) using each model applied to data sets with varying between- and within-patient variance ratios. The top row shows summaries of the within-patient data using estimated 50% concentration ellipses."
levs3 <- levels(Data$panel2)
levs3 <- sub(':fixed','', levs3)
levs3 <- sub(':BLUPs',' ', levs3)
Data$panel3 <- factor(sub(':fixed','', Data$panel2,), levels = levs3)
Datap$panel3 <- factor(sub(':BLUPs',' ', Data$panel2,), levels = levs3)

{   xyplot(y ~ x | SD_ratio * panel3, Data,
         groups = patient, type = 'n',
         drop.unused.levels = F,
         par.strip.text = list(cex = 0.4),
         scales = list(x = list(alternating = FALSE, cex = 0.7),
                       y = list(alternating =2, cex = 0.5)),
         xlab = list('dosage', cex = 0.7),
         between =
             list(y = c(0,0,0,0.25,0,0,0,.25)),
         ylab = list(
           paste0(
             paste0(rep(' ',18),  collapse = ''),
                  'fixed-effects',
             paste0(rep(' ',52), collapse = ''),
             'BLUPs',
             paste0(rep(' ',32), collapse = ''),
             'data'),
           cex = 0.7)) +
    glayer(panel.ellipse(..., center.pch = 16, center.cex = 0.5,
                         level = 0.5),
           panel.abline(a = 10, b = 1)) +
    xyplot(pred  ~ x | SD_ratio * panel2, Datap, type = 'l',
           groups = patient,
           drop.unused.levels = F)
} -> re.plot
useOuterStrips(re.plot)

```

Data sets with relatively low between-patient variance result in strong shrinkage of fixed-effects predictions and also of BLUPs towards the between-patient relationship between `y` and `x` in the model with an intercept and `x` as fixed-effect predictors, `~ 1 + x`. The inclusion of a contextual variable in the model corrects this problem.

Although the BLUPs fit the observed data more closely than predictions based on fixed effects alone, the slopes of within-patient BLUPs do not conform with the within-patient slopes for the `~ 1 + x` model in the two data sets with the smallest between-patient variances.

For data with a small between-patient variance, fixed-effects predictions for the `~ 1 + x` model have a slope that is close to the between-patient slope but provide better overall predictions than the fixed-effect predictions for data sets with larger between-subject variance.  With data whose between-patient variance is relatively large, predictions based on the model with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model.

@@

We will discover that case-based cross-validation prefers the `~ 1 + x` model to the `~ 1` model for the '5 / 2.5' data set, but cluster-based cross-validation prefers the latter model to the former. The situation is entirely reversed with the '8 / 2.5' data set.

@@

We next carry out case-based cross-validation, which, as we have explained, is based on both fixed and predicted random effects (i.e., BLUPs), and cluster-based cross-validation, which is based on fixed effects only. In order to reduce between-model random variability in comparisons of models on the same data set, we apply `cv()` to the list of models created by the `models()` function (introduced previously), performing cross-validation with the same folds for each model.

```{r echo=FALSE,include=FALSE}
library(cv) # unclear why it's necessary to reload cv
.opts <- options(warn = -2) 
```
```{r cross-validation,cache=TRUE,message=FALSE,include=FALSE}
model_lists <- lapply(fits, function(fitlist) do.call(models, fitlist))
cvs_cases <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]], k = 10, 
              data = split(Data, ~ SD_ratio)[[i]])
         })
cvs_clusters <-
  lapply(1:Nv,
         function(i){
           cv(model_lists[[i]],  k = 10, 
              data = split(Data, ~SD_ratio)[[i]], 
              clusterVariables = 'patient')
         })
```
The results for all data sets, using both cluster-based and case-based cross-validation, are assembled and shown in Figure \ref{fig:cross-validation-data-plot}.
```{r cross-validation-data,include=FALSE}
names(cvs_clusters) <- names(cvs_cases) <- SD_ratio 
dsummary <- expand.grid(SD_ratio_i = names(cvs_cases), model = names(cvs_cases[[1]]))
dsummary$cases <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_cases[[SD_ratio_i]][[model]][['CV crit']])
  })
dsummary$clusters <-
  sapply(1:nrow(dsummary), function(i){
    with(dsummary[i,], cvs_clusters[[SD_ratio_i]][[model]][['CV crit']])
  })
```
```{r cross-validation-data-plot,echo=FALSE}
#| out.width = "100%",
#| fig.height = 3.3,
#| fig.cap = "10-fold cluster- and case-based cross-validation comparing mixed models with random intercepts and various fixed effects."
xyplot(clusters + cases ~ model|SD_ratio_i, dsummary,
       auto.key = list(space = 'top', columns = 2, cex = 0.7), 
       type = 'b', ylab = list('CV criterion (MSE)', cex = .7),
       xlab = '',
       par.strip.text = list(cex = 0.7),
       layout= c(Nv, 1),
       par.settings =
         list(superpose.line=list(lty = c(1, 2), lwd = 1),
              superpose.symbol=list(cex = 0.7)),
       scales = list(y = list(log = TRUE, cex = 0.7), 
                     x = list(alternating = F, rot = 45, cex = 0.7))) 
```
In summary, when between-cluster variance is relatively large, the model `~ 1 + x`, with `x` alone and without the contextual mean of `x`, is assessed as fitting very poorly by cluster-based CV, but relatively much better by case-based CV. With all our examples, the model `~ 1 + x + xm`, which includes both `x` and its contextual mean, produces better results using both cluster-based and case-based CV.  These conclusions are consistent with our observations based on graphing predictions from the various models, and they illustrate the desirability of assessing mixed-effect models at different hierarchical levels.

@@

The transformed numeric variables are much better-behaved (cf., Figure \ref{fig:Auto-explore}, on page \pageref{fig:Auto-explore}, and Figure \ref{fig:Auto-transformed-scatterplot-matrix}):
```{r Auto-transformed-scatterplot-matrix}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Scatterplot matrix for the transformed numeric variables in the \\code{Auto} data"
scatterplotMatrix(~ log(mpg) + displacement + horsepower + weight 
                  + acceleration, 
                  smooth=list(spread=FALSE), data=A, pch=".")
```

And the partial relationships in the model fit to the transformed data are much more nearly linear (cf., Figure \ref{fig:Auto-working-model}, on page \pageref{fig:Auto-working-model}, and Figure \ref{fig:Auto-CR-plots-transformed}):
```{r Auto-CR-plots-transformed}
#| out.width = "60%",
#| fig.height = 5,
#| fig.cap = "Component+residual plots for the model fit to the transformed \\code{Auto} data"
crPlots(m, pch=".")
```

After transforming both the numeric predictors and the response, we proceed to use the \code{stepAIC()} function in the \pkg{MASS} package to perform predictor selection, employing the BIC model-selection criterion (by setting the `k` argument of `stepAIC()` to $\log n$):
```{r}
library("MASS")
m.step <- stepAIC(m, k=log(nrow(A)), trace=FALSE)
summary(m.step)
```

The selected model includes three of the numeric predictors, \code{horsepower}, \code{weight}, and \code{acceleration}, along with the factors \code{year} and \code{origin}. We can calculate the MSE for this model, but we expect that the result will be optimistic because we used the whole data to help specify the model:
```{r MSE-whole-selected-model}
mse(Auto$mpg, exp(fitted(m.step)))
```

This is considerably smaller than the MSE for the original working model:
```{r MSE-working-model}
mse(Auto$mpg, fitted(m.auto))
```

A perhaps subtle point is that we compute the MSE for the selected model on the original \code{mpg} response scale rather than the log scale, so as to make the selected model comparable to the working model.[^medAbsErr]

@@

compareFolds(cvs)
```

The other arguments to \code{cv()} are (see \code{?cv:::cv.function} for additional optional arguments and details):
